\documentclass[11pt]{article}
\usepackage[margin=1in]{geometry}
\usepackage{amsmath,amssymb}
\usepackage{booktabs}
\usepackage{graphicx}
\usepackage{hyperref}
\usepackage{natbib}
\usepackage{xcolor}

\title{DPO-Driven Preference Elicitation over Optimization Problems}
\author{Patrick Cooper}
\date{\today}

\begin{document}
\maketitle

% ---------------------------------------------------------------------------
\section{Introduction}

This report investigates whether Direct Preference Optimization (DPO) enables
a language model to efficiently surface user preferences through an active
elicitation loop over optimization problems.

The core question is: when an LLM conducts the full preference elicitation
cycle --- generating diagnostic queries, interpreting user responses, and
producing allocation recommendations --- does DPO fine-tuning improve the
quality and efficiency of this process compared to a base model?

We evaluate across three optimization domains:
\begin{enumerate}
  \item \textbf{Resource Strategy Game} -- a sequential allocation game with
    regime switching (4-channel variant A; 6-channel variant B).
  \item \textbf{Stock Backtesting} -- a multi-asset portfolio with five
    historical-calibrated asset classes.
  \item \textbf{Supply Chain Procurement} -- budget allocation across five
    suppliers with different cost, reliability, and disruption profiles.
\end{enumerate}

The system infers latent user preference parameters
(discount factor $\gamma$, risk aversion $\alpha$, loss aversion $\lambda$)
through structured binary-choice interactions. DPO training is applied in two
curriculum phases: Phase~1 (quality floor enforcement) and Phase~2
(type-conditioned alignment). The trained model is compared against the base
model and an analytical baseline (particle filter + EIG) to measure DPO's
contribution to elicitation effectiveness.

\subsection{Hypotheses}

\begin{description}
  \item[H1] DPO Phase~2 produces significantly higher-alignment recommendations
    than the base model across all environments.
  \item[H2] DPO Phase~1 reduces quality floor violations compared to the base
    model.
  \item[H3] The LLM-driven elicitation loop shows improving recommendation
    quality over successive rounds (convergence).
  \item[H4] The analytical baseline (EIG + particle filter) provides an upper
    bound on alignment; DPO narrows the gap relative to the base model.
\end{description}

The prior generalization results (cross-domain transfer, theta stability) are
retained as supporting evidence that the underlying preference framework is
sound, but the primary contribution is the DPO elicitation study.

% ---------------------------------------------------------------------------
\section{Methods}

\subsection{Environments}

All environments extend a shared \texttt{BaseEnvironment} interface and expose
the same observation layout (wealth vector, market-state matrix, round index),
action space (simplex over $K$ channels), and preference-aware API
(\texttt{quality\_score}, \texttt{check\_quality\_floor},
\texttt{get\_optimal\_action}).

Regime switching uses a $3 \times 3$ Markov transition matrix shared across
all domains (bull/bear/neutral for the game and stock; normal/disrupted/crisis
for supply chain).

\subsection{Elicitation: Two Paradigms}

\textbf{Analytical baseline.} A particle-filter posterior with EIG-maximizing
binary-choice queries; each domain has a \texttt{ScenarioLibraryBase} subclass
generating $\gamma$, $\alpha$, $\lambda$-discriminating scenarios. The optimal
recommendation is computed analytically via \texttt{get\_optimal\_action}.

\textbf{LLM-in-the-loop.} The language model drives the full cycle: given the
serialized environment state and interaction history, the model generates two
allocation options as a diagnostic query; a synthetic user responds via
softmax-rational choice; after $N$ rounds, the model produces a final
recommendation. Three LLM conditions are compared: base model (no DPO),
DPO Phase~1 (quality-only), and DPO Phase~2 (quality + alignment).

\subsection{DPO Training}

Training uses QLoRA (4-bit NF4, LoRA rank 16, $\alpha=32$) on
Qwen2.5-1.5B-Instruct. Phase~1 pairs have chosen = high quality score,
rejected = floor-violating. Phase~2 pairs are type-conditioned: chosen =
optimal for sampled user, rejected = optimal for a contrasting type.

\subsection{Statistical Tests}

All pairwise comparisons use the Wilcoxon signed-rank test (one-sided where
directional hypotheses are stated).  Effect sizes are reported as rank-biserial
$r$ with Cohen's $d$ as a secondary measure.  Multiple comparisons across the
full family of tests are corrected with the Holm--Bonferroni step-down
procedure.

% ---------------------------------------------------------------------------
\section{Experimental Setup}

\subsection{Study Conditions}

Four conditions evaluated on each of three environments (game, stock, supply
chain):
\begin{enumerate}
  \item \textbf{Analytical baseline} -- EIG query selection, particle filter
    posterior, closed-form optimal recommendation.
  \item \textbf{Base LLM} -- Qwen2.5-1.5B-Instruct generating queries and
    recommendations without DPO.
  \item \textbf{DPO Phase 1} -- same architecture, trained on quality-only pairs.
  \item \textbf{DPO Phase 2} -- further trained on type-conditioned alignment
    pairs.
\end{enumerate}

\subsection{Parameters}

\begin{table}[h]
\centering
\begin{tabular}{ll}
\toprule
Parameter & Value \\
\midrule
Synthetic users per condition & 12 \\
Elicitation rounds & 8 \\
Particles & 500 \\
EIG samples & 200 \\
Scenarios per round & 30 \\
Stability sessions & 2 \\
\bottomrule
\end{tabular}
\caption{Experimental parameters (local run; CURC targets 2000 particles, 800 EIG, 30+ users).}
\label{tab:params}
\end{table}

% ---------------------------------------------------------------------------
\section{Results}

\subsection{H1: Within-Domain Personalization}

Within-domain personalization significantly outperforms generic (uniform)
recommendations for game-domain pairs.  Wilcoxon signed-rank tests reject
the null for game~A$\to$B ($p = 0.004$) and game~A$\to$stock ($p = 0.004$).
For supply-chain-targeted pairs, within-domain advantage is present but not
statistically significant at this sample size ($p > 0.05$, Holm-corrected).

\begin{table}[h]
\centering
\begin{tabular}{lcccc}
\toprule
Pair & Generic & Within & Cross & H1 $p$ \\
\midrule
Game A $\to$ Game B & 0.00 & 0.88 & 0.88 & 0.004 \\
Game A $\to$ Stock & 0.00 & 0.83 & 0.81 & 0.004 \\
Game A $\to$ Supply Chain & -- & 0.50 & 0.48 & 1.0 \\
Stock $\to$ Supply Chain & -- & 0.45 & 0.42 & 1.0 \\
\bottomrule
\end{tabular}
\caption{Mean alignment scores by condition (12 users, 8 rounds, 500 particles).}
\label{tab:h1_results}
\end{table}

\subsection{H2: Cross-Domain Transfer}

Cross-domain transfer significantly outperforms generic for game pairs
(game~A$\to$B: $p = 0.004$; game~A$\to$stock: $p = 0.004$).  Within-domain
does not significantly exceed cross-domain at this scale, consistent with
strong preference transfer when source and target share the same
$(\gamma, \alpha, \lambda)$ parameterization.

\subsection{H3: Parameter-Level Transfer}

Gamma ($\gamma$, time preference) shows the lowest cross-domain MAE across
all pairs, confirming H3.  Mean $\gamma$ MAE = 0.17, compared to $\alpha$
MAE = 0.42 and $\lambda$ MAE = 0.50.  The Spearman correlation between
structural similarity and total transfer MAE is $\rho = 0.80$, though with
only four pairs this is not statistically significant after correction
($p = 1.0$, Holm).

\begin{table}[h]
\centering
\begin{tabular}{lccc}
\toprule
Pair & $\gamma$ MAE & $\alpha$ MAE & $\lambda$ MAE \\
\midrule
Game A $\to$ Game B & 0.156 & 0.474 & 0.709 \\
Game A $\to$ Stock & 0.176 & 0.391 & 0.512 \\
Game A $\to$ Supply Chain & 0.164 & 0.369 & 0.466 \\
Stock $\to$ Supply Chain & 0.180 & 0.446 & 0.330 \\
\bottomrule
\end{tabular}
\caption{Per-parameter cross-domain MAE.  $\gamma$ consistently lowest.}
\label{tab:h3_results}
\end{table}

\subsection{H4: Active vs.\ Random Elicitation}

At local scale (500 particles, 200 EIG samples, 5 users per domain),
active elicitation does not significantly outperform random querying.
Error reduction percentages: game = 2.7\%, stock = $-0.4$\%,
supply chain = $-26.6$\%.  The 30\% efficiency target from README
Section~7.1 is not met; this is consistent with guidance notes that
higher particle and EIG counts on CURC are needed for reliable EIG
estimation.

\subsection{Theta Stability}

\begin{table}[h]
\centering
\begin{tabular}{lcccccc}
\toprule
 & \multicolumn{3}{c}{ICC(2,1)} & \multicolumn{3}{c}{Pearson} \\
Domain & $\gamma$ & $\alpha$ & $\lambda$ & $\gamma$ & $\alpha$ & $\lambda$ \\
\midrule
Game A & 0.987 & 0.986 & 0.962 & 0.986 & 0.985 & 0.999 \\
Stock & 0.638 & 0.249 & $-0.137$ & 0.690 & 0.271 & $-0.125$ \\
Supply Chain & 0.266 & 0.107 & $-0.202$ & 0.262 & 0.101 & $-0.261$ \\
\bottomrule
\end{tabular}
\caption{Test-retest reliability (12 users, 2 sessions).  Game domain shows
excellent stability; stock and supply chain are lower, consistent with
higher per-period variance in those environments.}
\label{tab:stability}
\end{table}

% ---------------------------------------------------------------------------
\section{Discussion}

\subsection{Limitations}

\begin{itemize}
  \item The number of directed domain pairs (four) limits the power of the
    H3 similarity--transfer correlation; bootstrap confidence intervals
    supplement the Spearman test.
  \item All users are synthetic (softmax-rational).  Transfer to real human
    decision noise remains untested.
  \item Supply chain and stock environments use simplified regime models;
    real-world fidelity is limited.
\end{itemize}

\subsection{Structural Similarity Interpretation}

The structural similarity scores ($0.40$--$0.95$) are assigned based on
shared architecture (same $\gamma$/$\alpha$/$\lambda$ parameterization,
regime switching) and domain semantics.  Future work could learn similarity
from data.

% ---------------------------------------------------------------------------
\appendix
\section{Full Metric Tables}

\textit{Exported from \texttt{outputs/generalization/results\_bundle.json}.}

% Placeholder tables
% \input{tables/h1_results.tex}
% \input{tables/h2_results.tex}
% \input{tables/transfer_matrix.tex}
% \input{tables/stability.tex}

\end{document}
